The Matt Trick in Stan

### 2016-09-06 _knitr-posts/2015-02-02-stan-matt-trick.Rmd
Play with Stan Matt trick draft

```{r}
set.seed(123)
N <- 50 # number of observations per group
Nj <- 16 # number of groups
sigma_j <- 2.0 # standard deviation of intercepts across groups
d <- data.frame(j = 1:Nj, jname = letters[1:Nj], b0_j = rnorm(Nj, 0, sigma_j))
sigma <- 1.1 # residual standard deviation
b0 <- 0.3 # intercept
b1 <- 0.5 # slope
x <- lapply(seq_len(Nj), function(x) rnorm(n = N)) # predictor
dat <- list()
for(j in d$j) {
  dat[[j]] <- data.frame(j = j, jname = d$jname[j], x = x[[j]],
    y = rnorm(N, mean = b0 + x[[j]] * b1 + d$b0_j[j], sd = sigma))
}
dat <- as.data.frame(do.call("rbind", dat))
library("ggplot2")
ggplot(dat, aes(x, y)) + geom_point() + facet_wrap(~jname)
```

@@

```{r}
model <- "
data {
  int<lower=1> N;        // rows of data
  int<lower=1> Nj;      // number of groups
  int<lower=1,upper=Nj> j_id[N];  // group ids
  vector[N] x;       // predictor
  real y[N]; // response
}
parameters {
  real<lower=0> sigma;  // residual standard deviation
  real b0;              // intercept
  real b1;              // slope
  real<lower=0> sigma_j; // group-level standard deviation
  vector[Nj] b0_j;
}
transformed parameters {
  vector[N] yhat;
  for (i in 1:N) {
    yhat[i] <- b0_j[j_id[i]] + b1 * x[i];
  }
}
model {
  b0_j ~ normal(b0, sigma_j); 
  // priors:
  sigma ~ cauchy(0, 2.5);
  sigma_j ~ cauchy(0, 2.5);
  b0 ~ normal(0, 2.5);
  b1 ~ normal(0, 2.5);
  y ~ normal(yhat, sigma);
}
"
sm <- rstan::stan_model(model_code = model)
```

@@

```{r}
model <- "
data {
  int<lower=1> N;                 // rows of data
  int<lower=1> Nj;                // number of groups
  int<lower=1,upper=Nj> j_id[N];  // group ids
  vector[N] x;                    // predictor
  real y[N];                      // response
}
parameters {
  real<lower=0> sigma;            // residual standard deviation
  real b0;                        // intercept
  real b1;                        // slope
  real<lower=0> sigma_j;          // group-level standard deviation
  vector[Nj] e_b0_j;              // Matt Trick #1
}
transformed parameters {
  vector[N] yhat;
  vector[Nj] b0_j;                // Matt Trick #2
  b0_j <- b0 + sigma_j * e_b0_j;  // Matt Trick #3

  for (i in 1:N) {
    yhat[i] <- b0_j[j_id[i]] + b1 * x[i];
  }
}
model {
  e_b0_j ~ normal(0, 1);          // Matt Trick #4
  sigma ~ cauchy(0, 2.5);
  sigma_j ~ cauchy(0, 2.5);
  b0 ~ normal(0, 2.5);
  b1 ~ normal(0, 2.5);
  y ~ normal(yhat, sigma);
}
"
sm_matt <- rstan::stan_model(model_code = model)
```

@@

```{r}
model_data <- list(N = nrow(dat), y = dat$y, x = dat$x, Nj = max(dat$j), 
  j_id = dat$j)
pars <- c("b0", "b1", "sigma", "sigma_j")

st <- system.time({
  m <- sampling(sm, data = model_data, iter = 400, chains = 3, 
    pars = pars, seed = 999)
})
st_matt <- system.time({
  m_matt <- sampling(sm_matt, data = model_data, iter = 400, chains = 3, 
    pars = pars, seed = 999)
})
```

>>>>

```{r dat}
library("dplyr")
library("ggplot2")
library("rstan")
library("purrr")
# rstan_options(auto_write = TRUE)
# options(mc.cores = parallel::detectCores())
# options(mc.cores = 1)

sim_data <- function(N = 10, Nj = 10, sigma_j = 0.6, sigma = 1.1, 
  seed = 1, b0 = -0.3, b1 = 0.9, scale = FALSE) {
  set.seed(seed)
  # N <- 4 # number of observations per group
  # Nj <- 10 # number of groups
  # sigma_j <- 0.6 # standard deviation of intercepts across groups
  d <- tibble(j = 1:Nj, jname = seq_len(Nj), b0_j = rnorm(Nj, 0, sigma_j))
  # sigma <- 1.1 # residual standard deviation
  # b0 <- -0.3 # intercept
  # b1 <- 0.9 # slope
  x <- map(seq_len(Nj), function(x) runif(N, 0, 2)) # predictor
  dat <- list()
  for(j in d$j) {
    dat[[j]] <- data.frame(j = j, jname = d$jname[[j]], x = x[[j]],
      y = rnorm(N, mean = b0 + x[[j]] * b1 + d$b0_j[j], sd = sigma), 
      stringsAsFactors = FALSE)
  }
  dat <- bind_rows(dat)
  if (scale) dat <- mutate(dat, x = arm::rescale(x))
  dat
}
# ggplot(dat, aes(x, y)) + geom_point() + facet_wrap(~jname)
```

@@

```{r model}
model <- "
data {
  int<lower=1> N;        // rows of data
  int<lower=1> Nj;      // number of groups
  int<lower=1,upper=Nj> j_id[N];  // group ids
  vector[N] x;       // predictor
  real y[N]; // response
}
parameters {
  real<lower=0> sigma;  // residual standard deviation
  real b0;              // intercept
  real b1;              // slope
  real<lower=0> sigma_j; // group-level standard deviation
  vector[Nj] b0_j;
}
transformed parameters {
  vector[N] yhat;
  for (i in 1:N) {
    yhat[i] = b0_j[j_id[i]] + b1 * x[i];
  }
}
model {
  b0_j ~ normal(b0, sigma_j); 
  // priors:
  sigma ~ cauchy(0, 2.5);
  sigma_j ~ cauchy(0, 2.5);
  b0 ~ normal(0, 2.5);
  b1 ~ normal(0, 2.5);
  y ~ normal(yhat, sigma);
}
"
```

@@

```{r model_matt}
model_matt <- "
data {
  int<lower=1> N;                 // rows of data
  int<lower=1> Nj;                // number of groups
  int<lower=1,upper=Nj> j_id[N];  // group ids
  vector[N] x;                    // predictor
  real y[N];                      // response
}
parameters {
  real<lower=0> sigma;            // residual standard deviation
  real b0;                        // intercept
  real b1;                        // slope
  real<lower=0> sigma_j;          // group-level standard deviation
  vector[Nj] e_b0_j;              // Matt Trick #1
}
transformed parameters {
  vector[N] yhat;
  vector[Nj] b0_j;                // Matt Trick #2
  b0_j = b0 + sigma_j * e_b0_j;  // Matt Trick #3

  for (i in 1:N) {
    yhat[i] = b0_j[j_id[i]] + b1 * x[i];
  }
}
model {
  e_b0_j ~ normal(0, 1);          // Matt Trick #4
  sigma ~ cauchy(0, 2.5);
  sigma_j ~ cauchy(0, 2.5);
  b0 ~ normal(0, 2.5);
  b1 ~ normal(0, 2.5);
  y ~ normal(yhat, sigma);
}
"
```

@@

```{r fits, dependson=c("model", "model_matt", "dat"), cache=TRUE}
dat <- sim_data(N = 20, Nj = 8, seed = 999)
model_data <- list(N = nrow(dat), y = dat$y, x = dat$x, Nj = max(dat$j), 
  j_id = dat$j)
pars <- c("b0", "b1", "sigma", "sigma_j", "b0_j")

st <- system.time({
  m <- stan(model_code = model, data = model_data, iter = 600, 
    chains = 4, pars = pars, seed = 999)
})
st_matt <- system.time({
  m_matt <- stan(model_code = model_matt, data = model_data, iter = 600, 
    chains = 4, pars = pars, seed = 999)
})
```

```{r, dependson=fits}
# print(st)
# print(st_matt)
broom::tidyMCMC(m, rhat = TRUE, ess = TRUE) %>%
  mutate(ess_per_sec = round(ess / st[[1]], 0), time = st[[1]])
broom::tidyMCMC(m_matt, rhat = TRUE, ess = TRUE) %>%
  mutate(ess_per_sec = round(ess / st_matt[[1]], 0), time = st_matt[[1]])
```

```{r}
pars <- c("b0", "b1", "sigma", "sigma_j", "b0_j")

out <- plyr::ldply(c(2, 3, 5, 7, 10, 15, 20, 50), function(i) {
  dat <- sim_data(N = i, Nj = 10, seed = 2)
  model_data <- list(N = nrow(dat), y = dat$y, x = dat$x, Nj = max(dat$j), 
    j_id = dat$j)
  st <- system.time({
    m <- stan(model_code = model, data = model_data, iter = 300, 
      chains = 2, pars = pars, seed = 999)
  })
  st_matt <- system.time({
    m_matt <- stan(model_code = model_matt, data = model_data, iter = 300, 
      chains = 2, pars = pars, seed = 999)
  })
  b1 <- broom::tidyMCMC(m, rhat = TRUE, ess = TRUE) %>%
  mutate(ess_per_sec = round(ess / st[[1]], 0), time = st[[1]])
  b2 <- broom::tidyMCMC(m_matt, rhat = TRUE, ess = TRUE) %>%
  mutate(ess_per_sec = round(ess / st_matt[[1]], 0), time = st_matt[[1]])
  data.frame(m = min(b1$ess_per_sec), m_matt = min(b2$ess_per_sec), N = i,
    time = b1$time[1], time_matt = b2$time[1])
})
select(out, N, m, m_matt) %>% 
  reshape2::melt(id.vars = "N") %>%
  ggplot(aes(N, value, colour = variable)) + geom_line()
```
